Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/manuscript.qmd
Mega edits by Audrey

---
title: "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"
short-title: "Derogations and Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations:
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Did states misuse international legal emergency provisions during the COVID-19 pandemic to justify human rights abuse or did they follow international human rights law? Many governments restricted citizens' freedom of movement, association, and assembly during the crisis, raising questions about states' commitments to international human rights law. Some states used derogations to communicate temporary suspension of international legal provisions in a proportional and non-discriminatory manner, while others did not. We explore the dynamics of democratic backsliding and derogation use during the pandemic. We find that backsliding states were more likely to issue human rights treaty derogations. These derogations had mitigating effects once issued. Backsliding states that issued derogations were more likely to communicate restrictions and were less likely to issue abusive and discriminatory policy during the pandemic. Derogations helped temper abuse in states not experiencing backsliding. However, derogations did not always protect against abuse and media transparency in backsliding states. These results lend support to the use of flexibility mechanisms in international law and find that most states did not use emergency derogations to heighten human rights violations. The study contributes to the understanding of how international legal measures may help mitigate elements of democratic backsliding during times of crisis.
date: 2024-07-01
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
keywords:
  - human rights
  - international law
  - pandemic
  - derogation
  - democratic backsliding
bibliography: references.json
csl: csl/apa.csl
link-citations: true
---

@@

While research has explored democratic backsliding during the pandemic and government repression [@GrassePavlikMatfess:2021], its effect on human rights [@AdhikariKingMurdie:2024] and the behavior of backsliding states within international organizations (IOs) [@MeyerroseNooruddin:2023], we know less about how backsliding impacts engagement with international human rights treaties. This study extends research by exploring the dynamics between democratic backsliding, international legal behavior, and human rights. We ask: did backsliding states abuse states of emergency and international human rights law during the pandemic? While some scholars warned of the risk of emergency powers leading to abuse [@Scheinin:2020], we do not yet know if derogations increased the risk of human rights abuse during the pandemic.

First, we descriptively explore the phenomena of pandemic-era democratic backsliding and international human rights treaty participation. Combating the pandemic required extraordinary health measures; derogations in international law authorized states to temporarily suspend their international treaty commitments to provide them with flexibility while responding to this crisis. By declaring a state of emergency and formally derogating from international human rights treaties, states could acknowledge that their pandemic measures were temporary, necessary, and proportional, with an aim to restore normalcy as soon as possible. We first ask whether experiencing pandemic democratic backsliding impacted states' use of such derogations---were states experiencing democratic backsliding still abiding by their international legal obligations? How did the presence or absence of derogations impact human rights practices in derogating states, and was this substantially different from states not experiencing democratic backsliding? Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that backsliding states were more likely to issue human rights treaty derogations. Our findings also suggest that the derogations had some mitigation effects once issued. Backsliding states that derogated were less likely to issue abusive and discriminatory policies. Backsliding states not following international legal procedures were more likely to utilize discriminatory measures and were likely to engage in major violations of non-derogable rights such as right to life, freedom from torture, and freedom of thought, conscience and religion. Non-backsliding states that derogated also seemed to have mitigation effects across some measures, decreasing their risk of discriminatory policies. However, derogations did not always protect against abuse and media transparency in backsliding states. Overall, while the pandemic provided reasons for concerns for deepening autocratization in many countries, our results show that even in states where such democratic backsliding was unfolding, governments were not completely dismissive of the international dimensions of democracy, human rights, and rule of law. Backsliding states issuing derogations did not, overall, use that opportunity to systematically abuse human rights.

@@

Legal scholar Thomas Ginsburg recently issued a call for international institutions and international human rights law to "be engaged in the promotion, support, and disciplining of democracy." He argued that even though there is no "right to democracy" in international law, these actors are the best equipped to protect it against democratic backsliding [@Ginsburg:2021, p. 136]. In a similar vein, in response to global trends of democratic backsliding, some states have called for an increase in transparency in international human rights law. The Netherlands, for example, outlined a goal of states to "acknowledge one's own shortcomings" and the "challenges that they face in the realm of human rights and treaty compliance" [@MinistryofForeignAffairs:2023, p. 13]. In this paper, we explore the relationship between international human rights law engagement and democratic backsliding. Rather than looking to what determines and potentially prevents democratic backsliding, we seek to understand whether backsliding in turn shapes how states participate within the international human rights regime, and specifically whether the decline in democracy means a heightened and opportunistic abuse of international human rights law. From the literatures on international law, reputation, and democratic erosion, we draw several expectations about these dynamics.

@@

While there exists robust research on the relationship between regime type and international treaty ratification and compliance, we know much less about what democratic backsliding means for human rights law engagement. However, existing research on regime transitions, human rights compliance, and commitment are informative for our expectations. Regime transition matters along several important fronts. Regimes transitioning toward democracy, or democratizing, are more likely to use the international human rights regime as a means of legitimation [@Moravcsik:2000], including through treaty commitment [@Comstock:2021]. Though new democracies are more likely to commit to broad human rights conventions more quickly, they are also more hesitant to commit to human rights treaties with more demands [@DaiTokhi:2023]. Regimes transitioning away from democracy---or experiencing democratic backsliding---are found to repress human rights more. @Davenport:1999, for example, finds that the transition towards autocratization has a significant and positive impact on repressive behavior while democratization promotes more open governance. The strength of domestic democratic institutions, in particular, is important for these findings [@HillJones:2014]. @Meyerrose:2020 argues that if IOs promote some domestic institutions but not others, backsliding can occur when domestic executives are empowered above other institutions. In examining democratic backsliding and human rights more specifically, @AdhikariKingMurdie:2024 find strong statistical support that democratic backsliding harmed human rights. The authors found that the degree or intensity of backsliding along with the length of the backsliding period contributed to further chances of human rights violations. @Ginsburg:2019 suggests that international and regional courts may be a means for activists to resist the spread of national-level democratic backsliding.

@@

Bringing these literatures together, we expect that states that issue derogations are more willing and capable of communicating to domestic and international audiences about potential and real rights restrictions during the COVID-19 pandemic. Concerns about international reputation often shape behavior and are connected to the desire to obtain certain goals---these usually include some form of aid, trade, membership and holding office in IOs, and receiving higher rankings on performance indicators [@BushZetterberg:2021; @KelleySimmons:2019; @LevitskyWay:2010]. Generally, when countries seek cooperation from the international community, they seek external legitimacy [@Kelley:2012] and will consequently abide by legal and normative commitments to uphold human rights and rule of law. Public international commitments, such as compliance with the ICCPR, can thus be an important way for states, especially those backsliding, to signal their intent to the international community, to agree to respect these rights moving forward.

However, not all states are interested in protecting their international reputation. Despite the widespread acceptance of the liberal international order after the collapse of the Soviet Union, the contracting boundaries of this Western order in recent years has changed how states weigh the risks of violating liberal norms [@CooleyNexon:2020]. There has been an increase in the number of states contesting this order. Many have alternative patron states such as Russia and China, and they have little concern about protecting their reputations in front of Western states [@Benabdallah:2019; @HackeneschBader:2020]. These states may choose to avoid derogating simply for the purpose of signaling that they do not accept international human rights regimes.

We posit that backsliding without derogating signals that the national government does not value its international reputation or is not institutionally capable of communicating transparently about potential and real restrictions and abuses during the pandemic. In other words, we expect that we can learn about international signaling and state behavior by examining the dynamics of how backsliding states used international law during crisis. Examining behavior during the COVID-19 pandemic will provide insights into state signaling and intentions during the crisis.

@@

To test these expectations, this paper conducts a series of quantitative analyses followed by a qualitative examination of illustrative cases mapping on to each quadrant of @tbl-expectations. From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

```{r tbl-data-sources}
#| tbl-cap: Sources for outcomes and explanatory variables
#| include: true

tribble(
  ~Variable,                                                    ~Description,                                                   ~Source,
  "Pandemic Backsliding Index (PanBack)",                       "0–1; higher values represent greater risk of backsliding; ≥ 0.3 considered high risk",     "Varieties of Democracy PanDem",
  "ICCPR derogation",                                           "Binary indicator",                                             "UN Treaty Collection",
  "New and cumulative COVID deaths and cases",                  "Counts",                                                       "World Health Organization",
  "Rule of law index",                                          "0–1; higher values represent greater respect for rule of law", "Varieties of Democracy",
  "Emergency public health measures",                           "Binary indicator",                                             "Oxford COVID-19 Government Response Tracker",
  "Human rights and policy outcomes",                           "Ordered categories for the severity of abuses",                "Varieties of Democracy PanDem"
) |>
  tt(width = c(35, 45, 20)) |>
  group_tt(i = list(
    "Democratic backsliding" = 1,
    "Treaty activity" = 2,
    "Other variables" = 3
  ), indent = 0) |>
  style_tt(i = c(1, 3, 5), j = 1, bold = TRUE, background = "#e6e6e6", align = "l") |>
  style_tt(j = 1:3, align = "l") |>
  style_tt(
    bootstrap_class = "table table-sm"
  )
```

@@

In this section, we briefly examine four states as illustrative cases of varying derogation and democratic backsliding behavior to better unpack how democratic backsliding and derogation interaction (or lack thereof) may have influenced state behavior during the pandemic. Earlier in the paper we posited that transparency about restrictions and violation would vary across states based on derogation and backsliding behavior. @tbl-mini-cases situates the cases across backsliding and derogation status. Most states did not issue derogations and did not experience high levels of backsliding. All of the states that experienced high risk of democratic backsliding and issued derogations were from the Central American or South America regions. The states that issued derogations but did not experience a high risk of democratic backsliding were from a mix of regions including South America, Central America, Asia, Eastern Europe, and Africa. We selected cases that depict both regional variation and low/high risk of backsliding. Including extreme cases should be an "easy test" of the influence of backsliding on human rights violations and other behaviors if democratic backsliding has an impact. In addition to the ICCPR, states also derogated from regional human rights treaties---the American Convention on Human Rights (ACHR) and the European Convention on Human Rights (ECHR) [@IstrefiHumburg:2020]. Therefore, the mini-cases below examine derogations to ICCPR, as well as the ACHR and ECHR, where applicable.

@@

## High risk of backsliding, Derogations present: Guatemala

On March 9, 2020, Guatemala became the first state to formally derogate from Articles 12 and 21 of the ICCPR, both of which concern the freedoms of movement, association, assembly and demonstration [@PeacefulAssemblyWorldwide:2021]. Guatemala also sent official notification of derogation to the Organization of American States (OAS) on March 23, 2020, stating the government's intention to derogate from Articles 15 and 22 of the ACHR [@MinisteriodeRelacionesExterioresdeGuatemala:2020]. Article 15 guarantees the right of peaceful assembly, while Article 22 protects the freedom of movement and residence. Counter to trends in many countries across the globe, Guatemala notified the ICCPR before it sent notifications of derogation to the OAS, demonstrating a clear commitment to the treaties and international law. In addition, from the 10 states that derogated from the ACHR, only 4, including Guatemala filed notification of derogations from the ICCPR [@IstrefiHumburg:2020].

President Alejandro Giammattei was sworn in as President of Guatemala shortly before the pandemic. Prior democratic governments failed to meet their mandate---in 2019, more than half of the population lived below the poverty line [@FreemanPerello:2023]. Corruption was another big concern---in January 2020, Transparency International ranked Guatemala as the fourth most corrupt country in the world. Holding office was often seen as lucrative not just for politicians but also for those financing these leaders [@FreemanPerello:2023]. Thus, while many domestic and international observers were concerned about backsliding during his government, the pandemic also exacerbated fears about the relationship between transparency and corruption.

During the pandemic, Guatemala maintained a low rate of contagion compared with many other Latin American countries---which the government used to bolster its popularity. The OHCHR even praised the work done by the Office of Indigenous Women and the Ministry of Education for increasing support for indigenous children's participation in primary education. Between 2020--2023, there was a 7% increase in enrollment (49% of which was among girls); moreover, the dropout rates decreased by 4% counter to global trends during the pandemic [@OHCHR:2023]. In January 2021, the Guatemalan government requested derogations from two additional articles of the ACHR (Articles 13 and 16), and extensions to their previous derogations from Articles 15 and 22. Article 16 of the ACHR---like Article 21 of the ICCPR---protects the freedom of association, while Article 13 protects the freedom of thought and expression. Guatemala's derogation from articles regarding freedom of thought and expression has been seen by some as a concerning attempt to silence media criticisms of the government's handling of the pandemic [@TheGlobalStateofDemocracyInitiative:2021]. Human Rights Watch accused the administration of hindering journalists' access to public information [@Mercadal:2024]. This limited transparency also raised concerns about media freedom---though the government itself has not targeted any journalists, the media was shut out of various congressional sessions, making it unable to report accurately on the pandemic. Overall, though there were still causes for concern, Guatemala's derogations seemed to be lawful and generally proportionate, and the government engaged in fewer rights violations than expected. The Central American nation was quick to communicate its intentions to its treaty organizations, and in this case, the government's desire to communicate its intentions and maintain transparency may have provided a check on the rate of backsliding.

However, corruption was considered to be the main reason undermining efforts to fight the pandemic. Scholars noted that corruption during the pandemic "contributed to a regression of democracy" [@Mercadal:2024, p. 225]. Concerns about democratic backsliding ultimately did not fully materialize---in August 2023, Bernardo Arévalo, a centrist anti-corruption reformer, won Guatemala's presidential runoff by a wide margin after an electoral process that nearly saw Arévalo's party barred from competing.

## High risk of backsliding, no derogation: India

To illustrate the interaction of a high risk of democratic backsliding and the lack of derogations during the pandemic, we look at the case of India. According to Article 352 of India's constitution, India is only allowed to declare a state of emergency when its territory is threatened "by war or external aggression or armed rebellion" [@const-india-352]---not in the case of public health crises. Despite having an extremely deadly Delta wave of Covid-19 cases during the pandemic and the implementation of multiple measures to protect a sixth of the world's population, India did not derogate from any treaties. Without derogations, there is an absence of sunset clauses that typically ensure that there is an end to the measures a country implements during an emergency. Thus, many measures disregarded ICCPR provisions that should have been protected even in a state of emergency.

When India went into lockdown in March 2020, the government only provided a four-hour notice---this violated ICCPR Article 19, the right to seek and receive information, including early warnings of national measures like the lockdown. Further, the CESCR General Comment No. 14 says that "access to information concerning the main health problems in the community, including methods of preventing and controlling them" is also a guaranteed right [@AmnestyInternational:2020]. The lack of notice stranded a large number of migrant workers in cities, far from their homes in rural areas, with no transportation. Many of these workers died while trying to walk hundreds of miles back to their villages [@ChaudhryPrasad:2020].

As the pandemic progressed, the Bharatiya Janata Party (BJP)-led government used the pandemic measures to hasten backsliding. The government paid little heed to technical and scientific advice [@Mukherji:2020]. Instead, the BJP government used its institutional power to shut down dissent, especially from media, civil society, and lawyers. For instance, many measures disproportionately impacted journalists' ability to work [@icnl-covid]. While these measures may have been intended to prevent misinformation, they were used to suppress journalists and activists. The government also used restrictions to arbitrarily arrest and detain opponents to the regime---including those protesting the government and its Hindu nationalist policies [@YasirSchultz:2020]. Subsequent to their arrest, detainees had limited access to legal counsel, which led to their continued detention. Many measures also violated the right to privacy. Concerns over enhanced surveillance techniques arose after multiple leaks of personal information of infected peoples which has led to discrimination and even assault. These measures led to cases of Muslims being assaulted, harassed, and denied medical attention or spikes in caste-based discrimination and violence during the pandemic [@Ayyub:2020]. Ultimately, many scholars and policymakers argued that these measures were used to tighten both the government's grip on media as well as provide a justification to centralize power [@Mukherji:2020].

In The Global State of Democracy 2021 Report, India had the most violations among democracies experiencing backsliding. In the same year, Varieties of Democracy relegated India as an "electoral autocracy," CIVICUS coded India's civil society environment as "repressed," and Reporters Without Borders in its World Press Freedom Index ranked India 161 out of 180 countries [@Tripathi:2023]. Thus, over the course of the pandemic, India did not derogate from its treaty obligations, and its pandemic measures were further used to tighten both the government's grip on the media as a justification to centralize power. These developments occurred in the context of India continuing to maintain an independent relationship with Russia, despite pressure from the West. India repeatedly abstained from UN resolutions condemning Russia, and India's oil imports from Russia also increased after Western sanctions against the country [@Grossman:2022]. Thus, analysts have noted that while India is not abandoning the liberal international order, it has also ensured that by refusing to condemn Russia, it continues to receive tangible economic and security benefits from Russia [@Grossman:2022a].

## Low risk of backsliding, issued derogations: Armenia

Armenia declared a state of emergency in March 2020 and promptly derogated from both the ICCPR and ECHR. The declaration of a state of emergency in Armenia through Decree No. 298-N on March 16, 2020, resulted in the suspension of certain constitutional rights and freedoms, including freedom of movement and peaceful assembly [@icnl-covid]. The derogations were later extended in accordance with international law. Armenia also officially notified the Secretary General of the Council of Europe (COE) of possible derogations from the obligations of Armenia under the Convention [@CouncilofEurope:2020a]. On the 16th September 2020, Armenia withdrew all derogations and returned to full compliance with ICCPR [@PeacefulAssemblyWorldwide:2021].

Unlike countries experiencing a high risk of backsliding, the Armenian government worked with civil society representatives and the media to formulate pandemic measures pertaining to them [@CouncilofEurope:2020]. These recommendations were subsequently incorporated into a government decree adopted on 24 March 2020, revising the restrictions, which was welcomed also by the Organisation for Security and Cooperation in Europe (OSCE) Media Freedom Representative, as well as commended by the Armenian media [@CouncilofEurope:2020; @OSCE:2020]. Furthermore, in its State Reply to the COE on April 13, 2020, the government declared that the restrictions of media activities became void. The rationale was that the Armenian government was highly confident "in the information on COVID-19 provided by official sources among population and a responsible behavior of the media during this period" [@CouncilofEurope:2020]. Thus, unlike many other countries, media restrictions were lifted shortly after the state emergency was declared.

Armenia's derogations, and transparency not just with the treaty bodies, but also to the COE, as well as respect for media freedoms during this period can be explained, at least in part, by its increasing pivot to the West and desire to engage with the liberal international order. Russian-Armenian relations have been in decline, and there have been increasing discussions about Armenian desire to seek EU candidacy, breaking decades of affiliation with Russia [@Castillo:2024]. After repeated Russian passivity over Azerbaijan's offensives into Nagorno-Karabakh, Armenia froze relations with the Russian-led Collective Security Treaty Organization, organized military exercises with the US and expanded ties with democratic countries [@Kucera:2023]. In December 2023, the Armenian Foreign Minister hoped that Armenia would, "get as close to the European Union as the EU deems possible" [@Castillo:2024].

## Low risk of backsliding, no derogations: Hungary

Since taking office in 2010, Hungarian Prime Minister Viktor Orban implemented a number of constitutional and legal changes to consolidate his party's control over the country's institutions. In March 2020, in response to the pandemic, his government declared a national "State of Danger"---a special state of emergency [@icnl-covid]. Under this State of Danger, in addition to quarantining and social distancing regulations, and temporary closure of educational institutions, the government also increased police and military presence in the streets, border controls, and entry bans [@icnl-covid]. It also restricted data protection rights mandated by the General Data Protection Regulation (GDPR), an EU agreement that regulates information privacy. These restrictions allowed the government to use personal data of citizens without oversight [@MasseSzaboDenes:2020]. However, in implementing these measures, the government never notified the UN about its intent to derogate from the ICCPR.

The enforcement of these pandemic-related measures resulted in numerous violations of ICCPR and of derogation standards of non-discriminatory and proportional measures. In addition, many measures also restricted media freedom. Under one law, journalists who published "false" information about the pandemic or distorted government narratives would be punished with five years in jail [@icnl-covid]. The government also limited access to press conferences, only responded to media inquiries from pro-government outlets, and banned local health representatives from talking to the media [@InternationalCommissionofJurists:2022]. A 2022 report by the International Commission of Jurists on Hungary noted that, "By exercising emergency powers in order to justify the adoption of these measures, the government has failed to comply with or adequately consider international law standards with which such measures clearly conflict" [@InternationalCommissionofJurists:2022]. European Union Parliament lawmakers, in turn, demanded official punishment and denunciation of Hungary over some of these laws [@Cox:2020].

Hungary's brazen violation of not just the ICCPR, but also European regulations such as the GDPR can be reflective of its increasing rift from the European Union (EU) and its pivot towards Russia. Hungary buys billions of dollars in Russian oil and gas, despite many in the West ceasing to do so after Russia's invasion of Ukraine [@GavinJackGijs:2024]. Unlike other countries that voluntarily divested from Russian gas, Hungary even struck new deals with Moscow [@GavinJackGijs:2024]. Meanwhile, Orban has also criticized EU sanctions on Russia and blocked EU financial assistance for Ukraine [@Ridgwell:2024]. However, Russia is not the only alternative source of goods for Hungary---more recently, China has filled in this role. In 2023, Hungary was among the largest global recipients of Chinese Belt and Road Initiative investment to finance a high-speed railway from Budapest to Serbia [@Ridgwell:2024]. Thus, the presence of alternative patron states may have emboldened Hungary to violate not just international, but also European laws.

## Discussion

The four case illustrations show how derogations had a mitigating effect both in states experiencing democratic backsliding and those that did not. Derogating states in both cases had concerns about maintaining international reputation and legitimacy, and derogations helped them communicate their pandemic measures in a more or less transparent manner. However, states not as interested in protecting their international reputation, especially in the perspective of Western states, were less likely to use derogations as a means to communicate their intentions about pandemic measures. These states subsequently engaged in more abusive and discriminatory enforcement of such measures. The dynamics illustrated in these cases overall support the purpose of derogations in international law. However, state motivations for declaring derogations among states both experiencing backsliding or not are more complex and merits further investigation.

@@

In this paper we set out to explore the dynamics between international legal engagement via derogation and democratic backsliding. Given the global trend in democratic backsliding and concerns about the robustness of democratic institutions, we examinde how and why states used legal loopholes to limit their international human rights law obligations. We argued that the interaction between these two phenomena could provide insights into state behavior based, in part, on transparency around crises.

@@

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape, or mitigate, backsliding itself. Overall, this piece contributes to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic. Looking at both practices together helps us to understand how states operated in and signaled about their human rights during the COVID-19 pandemic.

>>>>

---
title: "Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic"
short-title: "Derogations and Backsliding"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations:
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Did states misuse international legal emergency provisions during the COVID-19 pandemic to justify human rights abuse or did they follow international human rights law? Many governments restricted citizens' freedom of movement, association, and assembly during the crisis, raising questions about states' commitments to international human rights law. Some states used derogations to communicate temporary suspension of international legal provisions in a proportional and non-discriminatory manner, while others did not. We explore the dynamics of democratic backsliding and derogation use during the pandemic. We find that backsliding states were more likely to issue human rights treaty derogations. These derogations had mitigating effects once issued. Backsliding states that issued derogations were more likely to communicate restrictions and were less likely to issue abusive and discriminatory policy during the pandemic. Derogations helped temper abuse in states not experiencing backsliding. However, derogations did not always protect against abuse and media transparency in backsliding states. These results lend support to the use of flexibility mechanisms in international law and find that most states did not use emergency derogations to heighten human rights violations. The study contributes to the understanding of how international legal measures may help mitigate elements of democratic backsliding during times of crisis.
date: 2024-11-24
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/lemon-lucifer>"
keywords:
  - human rights
  - international law
  - pandemic
  - derogation
  - democratic backsliding
bibliography: references.json
csl: csl/apa.csl
link-citations: true
---

@@

While research explores democratic backsliding during the pandemic and government repression [@GrassePavlikMatfess:2021], its effect on human rights [@AdhikariKingMurdie:2024] and the behavior of backsliding states within international organizations (IOs) [@MeyerroseNooruddin:2023], we know less about how backsliding impacts engagement with international human rights treaties. This study extends research by exploring the dynamics between democratic backsliding, international legal behavior, and human rights. We ask: did backsliding states abuse states of emergency and international human rights law during the pandemic? While some scholars warned of the risk of emergency powers leading to abuse [@Scheinin:2020], we do not yet know if derogations increased the risk of human rights abuse during the pandemic.

First, we descriptively explore the phenomena of pandemic-era democratic backsliding and international human rights treaty participation. Combating the pandemic required extraordinary health measures; derogations in international law authorized states to temporarily suspend their international treaty commitments to provide them with flexibility while responding to this crisis. By declaring a state of emergency and formally derogating from international human rights treaties, states could acknowledge that their pandemic measures were temporary, necessary, and proportional, with an aim to restore normalcy as soon as possible. We first ask whether experiencing pandemic democratic backsliding impacted states' use of derogations---were states experiencing democratic backsliding still abiding by their international legal obligations? How did the presence or absence of derogations impact human rights practices in derogating states, and was this substantially different from states not experiencing democratic backsliding? Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that backsliding states were more likely to issue human rights treaty derogations and that the derogations had some mitigation effects. Overall, while the pandemic provided reasons for concerns for deepening autocratization in many countries, our results show that even in states where such democratic backsliding was unfolding, governments were not completely dismissive of the international dimensions of democracy, human rights, and rule of law. Backsliding states issuing derogations did not, overall, use that opportunity to systematically abuse human rights.

@@

Did democratic backsliding shape how states participated within the international human rights regime, and specifically did the decline in democracy mean heightened, opportunistic abuse of international human rights law? From the literatures on international law, reputation, and democratic erosion, we draw several expectations about these dynamics. Namely, we emphasize the importance of transparency and legitimacy in the interplay between backsliding and derogations during the pandemic.

@@

Existing research on regime transitions, human rights compliance, and commitment are informative for our expectations about backsliding states and legal engagement. Regime transition matters along several important fronts. Regimes transitioning toward democracy, or democratizing, are more likely to use the international human rights regime as a means of legitimation [@Moravcsik:2000], including through treaty commitment [@Comstock:2021]. Though new democracies are more likely to commit to broad human rights conventions more quickly, they are also more hesitant to commit to human rights treaties with more demands [@DaiTokhi:2023]. Regimes transitioning away from democracy---or experiencing democratic backsliding---are found to repress human rights more. @Davenport:1999, for example, finds that the transition towards autocratization has a significant and positive impact on repressive behavior while democratization promotes more open governance. The strength of domestic democratic institutions, in particular, is important for these findings [@HillJones:2014]. @Meyerrose:2020 argues that if IOs promote some domestic institutions but not others, backsliding can occur when domestic executives are empowered above other institutions. In examining democratic backsliding and human rights more specifically, @AdhikariKingMurdie:2024 find strong statistical support that democratic backsliding harmed human rights. The authors found that the degree or intensity of backsliding along with the length of the backsliding period contributed to further chances of human rights violations. @Ginsburg:2019 suggests that international and regional courts may be a means for activists to resist the spread of national-level democratic backsliding. @MechkovaLuhrmannLindberg:2017 caution against interpreting democratic backsliding as too extensive of a trend, noting that most liberal and electoral democracies that declined were unlikely "to collapse so fully that it becomes an autocracy" (165). Rather, democratic character remains, at least in part, in these states experiencing democratic backsliding wherein the "normative power of democracy remains relatively strong" (168).

@@

Bringing these literatures together, we expect that states that issue derogations are more willing and capable of communicating to domestic and international audiences about potential and real rights restrictions during the COVID-19 pandemic. Concerns about international reputation often shape behavior and are connected to the desire to obtain certain goals---these usually include some form of aid, trade, membership and holding office in IOs, and receiving higher rankings on performance indicators [@BushZetterberg:2021; @KelleySimmons:2019; @LevitskyWay:2010]. Generally, when countries seek cooperation from the international community, they seek external legitimacy [@Kelley:2012] and will consequently abide by legal and normative commitments to uphold human rights and rule of law.

Public international commitments, such as compliance with the ICCPR, can thus be an important way for states, especially those backsliding, to signal their intent to the international community, to agree to respect these rights moving forward. Human rights compliance can be a costly signal for the costs associated with both reaching human rights standards and the threat of penalty if the standards are not met [@Moore:2003]. Human rights compliance and legal engagement provides information about states to others in the international community. More powerful or prominent states may feel the need to signal less given the information readily available about their interests and behavior [@Moore:2003, 890]. However, states with less information available about their interests, commitment, and behavior can use human rights treaty signaling to convey that information to a broad, global audience.

States in regime transitions that still have elements of democracy intact may be particularly interested in preserving their reputations and find that communicating information during a global crisis is one means of doing so. In this case, issuing derogations may become of particular interest to backsliding states as a means to communicate interests about human rights. A state’s reputation at the time of action can shape how impactful signaling around human rights can be and also incentivize signaling [@Guzman:2008, 8]. However, not all states are interested in protecting their international reputation [e.g., @Benabdallah:2019; @CooleyNexon:2020; @HackeneschBader:2020]. These states may choose to avoid derogating simply for the purpose of signaling that they do not accept international human rights regimes.

We posit that backsliding without derogating signals that the national government does not value its international reputation or is not institutionally capable of communicating transparently about potential and real restrictions and abuses during the pandemic. In other words, we expect that we can learn about international signaling and state behavior by examining the dynamics of how backsliding states used international law during a crisis. Examining behavior during the COVID-19 pandemic will provide insights into state signaling and intentions during the crisis.

@@

To test these expectations, we conduct a series of quantitative analyses followed by a qualitative examination of illustrative cases mapping on to each quadrant of @tbl-expectations. From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

@@

We briefly consider four states as illustrative cases of varying derogation and democratic backsliding behavior to better unpack how democratic backsliding and derogation interaction (or lack thereof) may have influenced state behavior during the pandemic. Earlier in the paper we posited that transparency about restrictions and violation would vary across states based on derogation and backsliding behavior. @tbl-mini-cases situates the cases across backsliding and derogation status. We selected cases that depict both regional variation and low/high risk of backsliding. Including extreme cases should be an "easy test" of the influence of backsliding on human rights violations and other behaviors if democratic backsliding has an impact. The four cases illustrate derogations' mitigating effect in states experiencing democratic backsliding and those that did not. Derogating states in both backsliding scenarios had concerns about maintaining international reputation and legitimacy (Armenia, Guatemala), and derogations helped them communicate their pandemic measures in a more or less transparent manner. However, states not as interested in protecting their international reputation, especially in the perspective of Western states, were less likely to use derogations as a means to communicate their intentions about pandemic measures (India, Hungary). These states subsequently engaged in more abusive and discriminatory enforcement of such measures. The dynamics illustrated in these cases overall support the purpose of derogations in international law. However, state motivations for declaring derogations among states both experiencing backsliding or not are more complex and merits further investigation. A full discussion of each of the cases is included in the appendix.

@@

@@

We set out to explore the dynamics between international legal engagement via derogation and democratic backsliding. Given the global trend in democratic backsliding and concerns about the robustness of democratic institutions, we examined how and why states used legal loopholes to limit their international human rights law obligations. We argued that the interaction between these two phenomena could provide insights into state behavior based, in part, on transparency around crises.

@@

What these findings point to is that democratic backsliding is a complex phenomenon---states experiencing backsliding impose discriminatory policies and abusive behaviors, but part of their institutional embeddedness remains invested, at least in part, in communicating some of these problems and maintaining connections with the international community around human rights. It is important to note that this study maps behavior that occurred but it does not fully theorize or test why backsliders were motivated to care about and submit derogations. We encourage future research on democratic backsliding and human rights to more fully examine backslider consideration of reputation, foreign policy during backsliding, and specifically international human rights law behavior during backsliding. Research might also track how continued engagement with international human rights law might shape, or mitigate, backsliding itself. Future research also would benefit the study of whether and how derogations might be used after a crisis has ended. Do states continue using crisis justification after a local/regional/global crisis is largely over? Overall, our findings contribute to the understanding of how neither derogation filing nor backsliding alone depict a complete picture of human rights and legal behaviors during the pandemic.
